Issues in developing multivariable molecular signatures for guiding clinical care decisions

### 2016-07-05 Documents/paper-revised-resub.Rmd
revisions

The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The principles discussed apply no matter what signature development method is used.  Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.  The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.  For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice). It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.  We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.  We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.

@@

To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study. Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 mutually independent features sampled from the standard normal distribution. This is the null case where no features are associated with $Y$. The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model. The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.

@@

```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. The true value of the AUC is 0.5. CV = Cross validation. \\label{fig1}"}
load("../Code/cvsim-result.RData")
cvres <- do.call("rbind", cvsims)
cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){

@@

names(labs) <- levels(cvlong$scen)

ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
  scale_x_discrete("Estimation Approach", labels = labs) + 
  ylab("Area Under the ROC Curve")
```

```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. The true value of the OR is 1.0. CV = Cross validation. \\label{fig2} "}

ggplot(subset(cvlong, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
  scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 1, 2, 4)) +
  ylab("Odds Ratio (log scale)")
```

@@

cvwide2 %>%
  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. CV = Cross validation. ")

@@

Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average. Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much. Investigators are tempted to use partial holdout estimates as it is more convenient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting performance estimates are still close to valid, as only half of the data are used to form the estimates.  However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.

>>>>

The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $\phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$. The principles discussed apply no matter what signature development method is used.  Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.  The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.  For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice). It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.  We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.  We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.

@@

To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study. Data were generated with $N = 200$ or 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and $p = 10, 100, 500,$ or $5000$ mutually independent features sampled from the standard normal distribution. This is the null case where no features are associated with $Y$. The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model. The 25 (10 in the case where $p = 10$) features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.

@@

```{r simset, include = FALSE}
load("../Code/cvsim-result.RData")
cvres <- do.call("rbind", cvsims)
cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){

@@

names(labs) <- levels(cvlong$scen)
```

```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment. In each replicate, there are $N$ observations and $p$ features. The true value of the AUC is 0.5. CV = Cross validation. \\label{fig1}", fig.height = 8.5}
source("../Code/new sims/newsims.R")

files <- grep("*.txt", list.files("../Code/new sims/"), value = TRUE)
results <- lapply(files, parse_sims)

alldat <- NULL
for(i in results){

  alldat <- rbind(alldat, i$datalong)

  names(labs) <- levels(alldat$scen)

ggplot(subset(alldat, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
    theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
    scale_x_discrete("Estimation Approach", labels = labs) +
    ylab("Area Under the ROC Curve") + facet_grid(p ~ N, labeller = "label_both")

```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment. In each replicate, there are $N$ observations and $p$ features. The true value of the OR is 1.0. CV = Cross validation. \\label{fig2} "}

alldat$value[alldat$stat == "OR"] <- exp(alldat$value[alldat$stat == "OR"])

ggplot(subset(alldat, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
    theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
    scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.1, .5, 1, 4, 10, 20)) +
    ylab("Odds Ratio (log scale)") + facet_grid(p ~ N, labeller = "label_both")

@@

cvwide2 %>%
  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate, there are 1000 observations and 500 features. CV = Cross validation. Additional results for different $N$ and $p$ are in the appendix. ")

@@

Not surprisingly, the resubstitution estimates are optimistically biased: for $N = 1000$ and $p = 500$ the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average. Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much. Investigators are tempted to use partial holdout estimates as it is more convenient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting performance estimates are still close to valid, as only half of the data are used to form the estimates.  However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.

It is evident in Figures \ref{fig1} and \ref{fig2} that the amount of bias tends to increase as the number of features increases. Even with only 10 features to select from, the resubstitution estimates are still biased. A larger sample size does not help with the bias, though it does improve precision of the performance estimates. With $N = 200$ and $p = 5000$, which is the situation closest to our real data example, we see that the resubstitution estimate indicates nearly perfect prediction performance, when in fact there is no signal in the data.

### 2016-07-05 Documents/paper-revised-resub.tex
revisions

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Michael C Sachs and Lisa M McShane},
            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{May 18, 2016}

@@

\begin{document}

\maketitle

@@

The focus of this paper is how to obtain a valid estimate of signature
performance as quantified by a metric \(phi\), that is, a good estimate
of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
apply no matter what signature development method is used. Performance
depends on the true signal in the data and the specific algorithm
\(\mathcal{F}\) used for development. The performance will reflect that
expected of signatures developed applying the prescribed methods to data
drawn from the distribution \(\mathcal{P}\). For signature deployment,
one would also want a particular specification of \(f\) for others to
use on independent data (e.g., in clinical practice). It is not trivial
to conduct this evaluation in a valid manner when there is a limited
data set available to both define a signature and assess its
performance. We illustrate the potential for bias to creep into the
performance estimation and review some strategies to avoid the pitfalls
that lead to these biases. We evaluate various strategies through a
simulation study and compare the performance of some of those strategies
using a real data exmaple involving the development of a signature for
use in making treatment decisions for patients with lung cancer.

@@

To illustrate some of the different properties of these estimates and
how they help to avoid overfitting, we conduct a limited simulation
study. Data were generated with 1000 observations, each with a binary
outcome \(Y\) with prevalence 0.3, and 500 mutually independent features
sampled from the standard normal distribution. This is the null case
where no features are associated with \(Y\). The signature development
procedure entails a feature selection step, in which each feature is
regressed against \(Y\) in a univariate logistic regression model. The
25 features with the smallest p-values are selected for inclusion in a
multivariable logistic regression model which defines the final
signature.

@@

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are 1000 observations and
500 features. The true value of the AUC is 0.5. CV = Cross validation.
\label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are 1000 observations and 500
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}

\begin{longtable}[c]{@{}lrrrrrr@{}}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a
dataset is used to both develop the signature and evaluate its
performance. The true value of the AUC is 0.5 and the true value of the
Log OR is 0.0. Estimates are based on 1000 replicates of the numerical
experiment. In each replicate, there are 1000 observations and 500
features. CV = Cross validation.}\tabularnewline
\toprule
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
Bias OR\tabularnewline
\midrule
\endfirsthead
\toprule
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
Bias OR\tabularnewline
\midrule
\endhead
Resubstitution & 0.72 & 0.01 & 0.22 & 1.33 & 0.12 & 1.33\tabularnewline
Partial CV & 0.68 & 0.01 & 0.18 & 1.02 & 0.12 & 1.02\tabularnewline
Partial Holdout & 0.67 & 0.02 & 0.17 & 0.97 & 0.19 & 0.97\tabularnewline
Partial Resubstitution & 0.65 & 0.02 & 0.15 & 0.88 & 0.12 &
0.88\tabularnewline
Pre-validation & 0.50 & 0.05 & 0.00 & -0.01 & 0.32 &
-0.01\tabularnewline
Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.25 & 0.00\tabularnewline
Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 &
0.00\tabularnewline
30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.24 & 0.00\tabularnewline
50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\tabularnewline
Bootstrap & 0.50 & 0.01 & 0.00 & 0.00 & 0.08 & 0.00\tabularnewline
\bottomrule
\end{longtable}

Not surprisingly, the resubstitution estimates are optimistically
biased: the naive resubstitution estimate of the AUC is 44\% larger than
it should be and the log OR estimate is over 2 times higher than it
should be, on average. Partial resubstitution, partial holdout, and
partial cross-validation estimates do not ameliorate the bias very much.
Investigators are tempted to use partial holdout estimates as it is more
convenient to not have to carry along a large number of feature
measurements into the modeling part of the signature development
process, and they may think that the resulting performance estimates are
still close to valid, as only half of the data are used to form the
estimates. However, here we see that these versions are still severely
biased and should not be reported as valid assessments of the
performance of biomarker signatures.

@@

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/examppresub-1.pdf}
\caption{Comparison of survival by gene expression based risk signature.
The left plot shows the resubstitution estimate, while the right plot
shows the pre-validated estimate. \label{fig2}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised_files/figure-latex/inter1-1.pdf}
\caption{Survival curves comparing the treatment effect by gene
expression signature risk group. The top set of plots shows the partial
resubstitution based signature and the bottom row shows the
pre-validated signature estimates. \label{fig3}}
\end{figure}

@@

\renewcommand\refname{References}
\bibliography{paper}

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
            pdfauthor={Michael C Sachs and Lisa M McShane},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{June 30, 2016}

@@

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

@@

The focus of this paper is how to obtain a valid estimate of signature
performance as quantified by a metric \(\phi\), that is, a good estimate
of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
apply no matter what signature development method is used. Performance
depends on the true signal in the data and the specific algorithm
\(\mathcal{F}\) used for development. The performance will reflect that
expected of signatures developed applying the prescribed methods to data
drawn from the distribution \(\mathcal{P}\). For signature deployment,
one would also want a particular specification of \(f\) for others to
use on independent data (e.g., in clinical practice). It is not trivial
to conduct this evaluation in a valid manner when there is a limited
data set available to both define a signature and assess its
performance. We illustrate the potential for bias to creep into the
performance estimation and review some strategies to avoid the pitfalls
that lead to these biases. We evaluate various strategies through a
simulation study and compare the performance of some of those strategies
using a real data exmaple involving the development of a signature for
use in making treatment decisions for patients with lung cancer.

@@

To illustrate some of the different properties of these estimates and
how they help to avoid overfitting, we conduct a limited simulation
study. Data were generated with \(N = 200\) or 1000 observations, each
with a binary outcome \(Y\) with prevalence 0.3, and
\(p = 10, 100, 500,\) or \(5000\) mutually independent features sampled
from the standard normal distribution. This is the null case where no
features are associated with \(Y\). The signature development procedure
entails a feature selection step, in which each feature is regressed
against \(Y\) in a univariate logistic regression model. The 25 (10 in
the case where \(p = 10\)) features with the smallest p-values are
selected for inclusion in a multivariable logistic regression model
which defines the final signature.

@@

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/cvsims-1.pdf}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) in the setting where a dataset is used to both
develop the signature and evaluate its performance. The violin plots
show mirrored density estimates for the AUC for 1000 replicates of the
numerical experiment. In each replicate, there are \(N\) observations
and \(p\) features. The true value of the AUC is 0.5. CV = Cross
validation. \label{fig1}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/cvsims2-1.pdf}
\caption{Comparison of different approaches to estimating the odds ratio
(OR) in the setting where a dataset is used to both develop the
signature and evaluate its performance. The violin plots show mirrored
density estimates for the log OR for 1000 replicates of the numerical
experiment. In each replicate, there are \(N\) observations and \(p\)
features. The true value of the OR is 1.0. CV = Cross validation.
\label{fig2}}
\end{figure}

\begin{longtable}[c]{@{}lrrrrrr@{}}
\caption{Comparison of different approaches to estimating the Area Under
the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a
dataset is used to both develop the signature and evaluate its
performance. The true value of the AUC is 0.5 and the true value of the
Log OR is 0.0. Estimates are based on 1000 replicates of the numerical
experiment. In each replicate, there are 1000 observations and 500
features. CV = Cross validation. Additional results for different \(N\)
and \(p\) are in the appendix.}\tabularnewline
\toprule
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
Bias OR\tabularnewline
\midrule
\endfirsthead
\toprule
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
Bias OR\tabularnewline
\midrule
\endhead
Resubstitution & 0.72 & 0.01 & 0.22 & 1.33 & 0.12 & 1.33\tabularnewline
Partial CV & 0.68 & 0.01 & 0.18 & 1.02 & 0.12 & 1.02\tabularnewline
Partial Holdout & 0.67 & 0.02 & 0.17 & 0.97 & 0.19 & 0.97\tabularnewline
Partial Resubstitution & 0.65 & 0.02 & 0.15 & 0.88 & 0.12 &
0.88\tabularnewline
Pre-validation & 0.50 & 0.05 & 0.00 & -0.01 & 0.32 &
-0.01\tabularnewline
Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.25 & 0.00\tabularnewline
Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 &
0.00\tabularnewline
30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.24 & 0.00\tabularnewline
50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\tabularnewline
Bootstrap & 0.50 & 0.01 & 0.00 & 0.00 & 0.08 & 0.00\tabularnewline
\bottomrule
\end{longtable}

Not surprisingly, the resubstitution estimates are optimistically
biased: for \(N = 1000\) and \(p = 500\) the naive resubstitution
estimate of the AUC is 44\% larger than it should be and the log OR
estimate is over 2 times higher than it should be, on average. Partial
resubstitution, partial holdout, and partial cross-validation estimates
do not ameliorate the bias very much. Investigators are tempted to use
partial holdout estimates as it is more convenient to not have to carry
along a large number of feature measurements into the modeling part of
the signature development process, and they may think that the resulting
performance estimates are still close to valid, as only half of the data
are used to form the estimates. However, here we see that these versions
are still severely biased and should not be reported as valid
assessments of the performance of biomarker signatures.

It is evident in Figures \ref{fig1} and \ref{fig2} that the amount of
bias tends to increase as the number of features increases. Even with
only 10 features to select from, the resubstitution estimates are still
biased. A larger sample size does not help with the bias, though it does
improve precision of the performance estimates. With \(N = 200\) and
\(p = 5000\), which is the situation closest to our real data example,
we see that the resubstitution estimate indicates nearly perfect
prediction performance, when in fact there is no signal in the data.

@@

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/examppresub-1.pdf}
\caption{Comparison of survival by gene expression based risk signature.
The left plot shows the resubstitution estimate, while the right plot
shows the pre-validated estimate. \label{fig2}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{paper-revised-resub_files/figure-latex/inter1-1.pdf}
\caption{Survival curves comparing the treatment effect by gene
expression signature risk group. The top set of plots shows the partial
resubstitution based signature and the bottom row shows the
pre-validated signature estimates. \label{fig3}}
\end{figure}

@@

\renewcommand\refname{References}
\bibliography{paper.bib}
